---
title: "METT SEM"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE)
```

## Summary 

We analyzed causal models of the METT data. First we built a full model based on hypothesized causal relationships, then we tested the sufficiency of its fit and modified the model to get an adequate fit, finally we tested three nested sub-models against the full modified model. 


```{r include = FALSE}

library(tidyverse)
library(corrplot)
library(Hmisc)
library(lavaan)
library(dagitty)
```


```{r}
dat <- read.csv("../METT/merged-mett.csv")

datm <- as.matrix(dat[,c(26,3:25,27)])
datm[datm>3] <- NA

cors <- rcorr(datm,
              type = "spearman")
corsig <- cors$r 
corsig[cors$P > 0.005] <- 0




dat2 <- dat %>%
  select(c("X14_Staff_Training",
           "X12_Staff_Numbers",
    "X15_Current_Budget",
           "X16_Security_of_Budget",
           "X20_Education_and_Awareness_Program",
           "X17_Management_of_Budget",
           "X23_Local_Communities",
           "X24_Visitor_Facilities",
           "X25_Commercial_Tourism",
           "X26_Fees"))
irm <- which(dat2>3, arr.ind = TRUE)
dat3 <- dat2[-unique(irm[,1]),] %>%
  mutate(across(starts_with("X"), ordered))
dat3$X26_Fees <- as.numeric(as.character(dat3$X26_Fees))

```

## Correlations

```{r fig.width=8, fig.height=8}
corrplot(corsig, method = "circle",
         tl.cex = 0.65,
         tl.srt = 45,
         diag = FALSE,
         type = "lower",
         order = "original")
```

**Figure S4** Spearman rank correlations among variables included in the METT database.

**Table S2** METT variables used in the analysis 

```{r}
knitr::kable(matrix(names(dat2), ncol = 1))
```


# Model fitting 
Note that in specifications below, variables starting with X are observed, variables in lower case are latent. 

## Full model 

Lavaan specification of the full model. 

```{r echo = TRUE}
model1 <- '
  budget =~ X15_Current_Budget + X16_Security_of_Budget + X17_Management_of_Budget
  budget ~ X26_Fees
  X15_Current_Budget ~~ X16_Security_of_Budget
  staff =~ X12_Staff_Numbers + X14_Staff_Training
  staff ~~ budget
  activities ~ budget + staff
  activities =~ X24_Visitor_Facilities + X25_Commercial_Tourism + X20_Education_and_Awareness_Program + X23_Local_Communities 
'
```

```{r}
s1 <- sem(model1, dat3, std.lv = TRUE)
```
```{r, eval = FALSE}
x <- lavaanToGraph(s1)
coordinates(x) <- list(
  x = c(budget = 0, engagement = 1, participation = 2,
        X15_Current_Budget = -1,
        X16_Security_of_Budget = -1,
        X17_Management_of_Budget = -1,
        X26_Fees = 0,
        X14_Staff_Training = 0.5,
        X12_Staff_Numbers = 1.5,
        X20_Education_and_Awareness_Program = 1,
        X24_Visitor_Facilities = 1,
        X23_Local_Communities = 3,
        X25_Commercial_Tourism = 3.3),
  y = c(budget = 0, engagement = 0, participation = 0,
        X15_Current_Budget = 1,
        X16_Security_of_Budget = 0,
        X17_Management_of_Budget = -1,
        X26_Fees = 1,
        X14_Staff_Training = -0.5,
        X12_Staff_Numbers = -0.5,
        X20_Education_and_Awareness_Program = 1,
        X24_Visitor_Facilities = -1,
        X23_Local_Communities = 1,
        X25_Commercial_Tourism = 0)
)
par(cex = 0.5)
plot(x,
     abbreviate.names = FALSE,
     show.coefficients = TRUE)

```


Chi-square statistic and p-value for model 1. 

```{r}
fitMeasures(s1, c("chisq","pvalue"))
```
Significant, indicating correlations in the data not explained by the model. 

Look at modification indices. High values of 'mi' indicate important correlations. 

```{r}
x <- modindices(s1)[,1:5]
x <- x[order(x$mi, decreasing = TRUE),]
x[1:10,]
```
We will add the unexplained correlation between staff and fees first as it is the most plausible correlation with a high modification index.  

### Model 1b 

As for the full model but with fees affecting staff

```{r echo = TRUE}
model1b <- '
  budget =~ X15_Current_Budget + X16_Security_of_Budget + X17_Management_of_Budget
  budget ~ X26_Fees
  X15_Current_Budget ~~ X16_Security_of_Budget
  staff =~ X12_Staff_Numbers + X14_Staff_Training
  staff ~ X26_Fees
  staff ~~ budget
  activities ~ budget + staff
  activities =~ X24_Visitor_Facilities + X25_Commercial_Tourism + X20_Education_and_Awareness_Program + X23_Local_Communities 
'
```

```{r}
s1b <- sem(model1b, dat3, std.lv = TRUE)
```
```{r, eval = FALSE}
x <- lavaanToGraph(s1b)

par(cex = 0.5)
plot(x,
     abbreviate.names = FALSE,
     show.coefficients = TRUE)

```

```{r}
fitMeasures(s1b, c("chisq","pvalue"))
```

Significant, indicating correlations in the data not explained by the model. 

Look at modification indices. High values of 'mi' indicate important correlations. 

```{r}
x <- modindices(s1b)[,1:5]
x <- x[order(x$mi, decreasing = TRUE),]
x[1:10,]
```
We will add the unexplained correlation between visitor facilities and commercial tourism as then next most plausible effect. 

### Model 1c

As for model 1b but with correlation between visitor facilities and commercial tourism. 

```{r echo = TRUE}
model1c <- '
  budget =~ X15_Current_Budget + X16_Security_of_Budget + X17_Management_of_Budget
  budget ~ X26_Fees
  X15_Current_Budget ~~ X16_Security_of_Budget
  staff =~ X12_Staff_Numbers + X14_Staff_Training
  staff ~ X26_Fees
  staff ~~ budget
  activities ~ budget + staff
  activities =~ X24_Visitor_Facilities + X25_Commercial_Tourism + X20_Education_and_Awareness_Program + X23_Local_Communities 
  X24_Visitor_Facilities~~X25_Commercial_Tourism
'
```

```{r}
s1c <- sem(model1c, dat3, std.lv = TRUE)
```
```{r}
x <- lavaanToGraph(s1c)

par(cex = 0.5)
plot(x,
     abbreviate.names = FALSE,
     show.coefficients = TRUE)

```

**Figure S5** The modified full model with path coefficients. 

```{r}
fitMeasures(s1c, c("chisq","pvalue"))
```

This model fits has p>0.05 indicating an acceptable fit. 

Polychoric correlatoins
```{r}
lavCor(s1c, ordered = TRUE, group = NULL, output = "cor")
```





```{r, include=FALSE}
### Save parameters
ss1c <- summary(s1c)
x <- ss1c$PE %>%
  filter(!(lhs == rhs)) %>%
  filter(!(op == "|")) %>%
  filter(!(op == "~1")) %>%
  mutate(est = signif(est, 3),
         se = signif(se, 3),
         z = signif(z, 3),
         pvalue = round(pvalue,3)) %>%
  select(LHS = lhs, RHS = rhs, `Path coefficient` = est, `S.E.` = se)
# write.csv(x, "params-mod1c.csv", row.names = FALSE)
```

**Table S3** Path coefficients and their approximate standard errors for the final version of model 1. LHS: Term on the left-hand side of the equation, RHS: term on the right-hand side of the equation. Variables prepended with ‘X’ are observed variables, numbers refer to the relevant METT survey question.  

## Predictions given entry fee input

```{r}
summary(s1c)
```


```{r}
knitr::kable(x)
```
## Testing nested sub-models  
 

### Staff capacity supports outreach activities 

```{r echo = FALSE}
model1c_3 <- '
  budget =~ X15_Current_Budget + X16_Security_of_Budget + X17_Management_of_Budget
  budget ~ X26_Fees
  X15_Current_Budget ~~ X16_Security_of_Budget
  staff =~ X12_Staff_Numbers + X14_Staff_Training
  staff ~ X26_Fees
  staff ~~ budget
  activities ~ budget 
  activities ~~ 0*staff
  activities =~ X24_Visitor_Facilities + X25_Commercial_Tourism + X20_Education_and_Awareness_Program + X23_Local_Communities 
  X24_Visitor_Facilities~~X25_Commercial_Tourism
'
s1c_3 <- sem(model1c_3, dat3, std.lv = TRUE)
anova(s1c_3, s1c)
```

### Budget health supports outreach activities 

```{r echo = FALSE}
model1c_4 <- '
  budget =~ X15_Current_Budget + X16_Security_of_Budget + X17_Management_of_Budget
  budget ~ X26_Fees
  X15_Current_Budget ~~ X16_Security_of_Budget
  staff =~ X12_Staff_Numbers + X14_Staff_Training
  staff ~ X26_Fees
  staff ~~ budget
  activities ~ staff
  activities ~~ 0*budget
  activities =~ X24_Visitor_Facilities + X25_Commercial_Tourism + X20_Education_and_Awareness_Program + X23_Local_Communities 
  X24_Visitor_Facilities~~X25_Commercial_Tourism
'
s1c_4 <- sem(model1c_4, dat3, std.lv = TRUE)
anova(s1c_4, s1c)

```

```{r}
x <- lavaanToGraph(s1c_4)

par(cex = 0.5)
plot(x,
     abbreviate.names = FALSE,
     show.coefficients = TRUE)

```  

**Figure S6** The nested sub-model that lacks effect of budget health on outreach activities.  

### Entry fees were a driver of budget health. 

```{r echo = FALSE}
model1c_5 <- '
  budget =~ X15_Current_Budget + X16_Security_of_Budget + X17_Management_of_Budget
  X15_Current_Budget ~~ X16_Security_of_Budget
  staff =~ X12_Staff_Numbers + X14_Staff_Training
  staff ~ X26_Fees
  staff ~~ budget
  activities ~ budget + staff
  activities =~ X24_Visitor_Facilities + X25_Commercial_Tourism + X20_Education_and_Awareness_Program + X23_Local_Communities 
  X24_Visitor_Facilities~~X25_Commercial_Tourism
'
s1c_5 <- sem(model1c_5, dat3, std.lv = TRUE)
anova(s1c_5, s1c)
```
